Creating Something Different: Similarity, Contrast, and Representativeness in Categorization

### 2024-02-16 Manuscripts/cbb2023/sn-article.tex
from overleaf

%%\documentclass[sn-basic]{sn-jnl}% Basic Springer Nature Reference Style/Chemistry Reference Style
%%\documentclass[sn-mathphys]{sn-jnl}% Math and Physical Sciences Reference Style
%%\documentclass[sn-aps]{sn-jnl}% American Physical Society (APS) Reference Style
%%\documentclass[sn-vancouver]{sn-jnl}% Vancouver Reference Style
\documentclass[sn-apa]{sn-jnl}% APA Reference Style
%%\documentclass[sn-chicago]{sn-jnl}% Chicago-based Humanities Reference Style
%%\documentclass[sn-standardnature]{sn-jnl}% Standard Nature Portfolio Reference Style
%%\documentclass[default]{sn-jnl}% Default
%%\documentclass[default,iicol]{sn-jnl}% Default with double column layout

%%%% Standard Packages
%%<additional latex packages if required can be included here>
%%%%

@@

%TODO: CONVERT FINAL PGFS TO STANDARD FORMAT
\pgfplotsset{compat=1.6}
\newcommand\inputpgf[2]{{
\let\pgfimageWithoutPath\pgfimage
\renewcommand{\pgfimage}[2][]{\pgfimageWithoutPath[##1]{#1/##2}}
\input{#1/#2}
}}

@@

\title[Article Title]{Article Title}

@@

\author*[1,2]{\fnm{First} \sur{Author}}\email{iauthor@gmail.com}

\author[2,3]{\fnm{Second} \sur{Author}}\email{iiauthor@gmail.com}
\equalcont{These authors contributed equally to this work.}

\author[1,2]{\fnm{Third} \sur{Author}}\email{iiiauthor@gmail.com}
\equalcont{These authors contributed equally to this work.}

\affil*[1]{\orgdiv{Department}, \orgname{Organization}, \orgaddress{\street{Street}, \city{City}, \postcode{100190}, \state{State}, \country{Country}}}

\affil[2]{\orgdiv{Department}, \orgname{Organization}, \orgaddress{\street{Street}, \city{City}, \postcode{10587}, \state{State}, \country{Country}}}

\affil[3]{\orgdiv{Department}, \orgname{Organization}, \orgaddress{\street{Street}, \city{City}, \postcode{610101}, \state{State}, \country{Country}}}

@@

\abstract{The abstract serves both as a general introduction to the topic and as a brief, non-technical summary of the main results and their implications. Authors are advised to check the author instructions for the journal they are submitting to for word limits and if structural elements like subheadings, citations, or equations are permitted.}

@@

\keywords{keyword1, Keyword2, Keyword3, Keyword4}

@@

\section{Introduction}\label{sec1}

The Introduction section, of referenced text \cite{bib1} expands on the background of the work (some overlap with the Abstract is acceptable). The introduction should not include subheadings.

Springer Nature does not impose a strict layout as standard however authors are advised to check the individual requirements for the journal they are planning to submit to as there may be journal-level preferences. When preparing your text please also be aware that some stylistic choices are not supported in full text XML (publication version), including coloured font. These will not be replicated in the typeset article if it is accepted.

\section{Results}\label{sec2}

Sample body text. Sample body text. Sample body text. Sample body text. Sample body text. Sample body text. Sample body text. Sample body text.

\section{This is an example for first level head---section head}\label{sec3}

\subsection{This is an example for second level head---subsection head}\label{subsec2}

\subsubsection{This is an example for third level head---subsubsection head}\label{subsubsec2}

Sample body text. Sample body text. Sample body text. Sample body text. Sample body text. Sample body text. Sample body text. Sample body text.

\section{Equations}\label{sec4}

Equations in \LaTeX\ can either be inline or on-a-line by itself (``display equations''). For
inline equations use the \verb+$...$+ commands. E.g.: The equation
$H\psi = E \psi$ is written via the command \verb+$H \psi = E \psi$+.

For display equations (with auto generated equation numbers)
one can use the equation or align environments:
\begin{equation}
\|\tilde{X}(k)\|^2 \leq\frac{\sum\limits_{i=1}^{p}\left\|\tilde{Y}_i(k)\right\|^2+\sum\limits_{j=1}^{q}\left\|\tilde{Z}_j(k)\right\|^2 }{p+q}.\label{eq1}
\end{equation}
where,
\begin{align}
D_\mu &=  \partial_\mu - ig \frac{\lambda^a}{2} A^a_\mu \nonumber \\
F^a_{\mu\nu} &= \partial_\mu A^a_\nu - \partial_\nu A^a_\mu + g f^{abc} A^b_\mu A^a_\nu \label{eq2}
\end{align}
Notice the use of \verb+\nonumber+ in the align environment at the end
of each line, except the last, so as not to produce equation numbers on
lines where no equation numbers are required. The \verb+\label{}+ command
should only be used at the last line of an align environment where
\verb+\nonumber+ is not used.
\begin{equation}
Y_\infty = \left( \frac{m}{\textrm{GeV}} \right)^{-3}
    \left[ 1 + \frac{3 \ln(m/\textrm{GeV})}{15}
    + \frac{\ln(c_2/5)}{15} \right]
\end{equation}
The class file also supports the use of \verb+\mathbb{}+, \verb+\mathscr{}+ and
\verb+\mathcal{}+ commands. As such \verb+\mathbb{R}+, \verb+\mathscr{R}+
and \verb+\mathcal{R}+ produces $\mathbb{R}$, $\mathscr{R}$ and $\mathcal{R}$
respectively (refer Subsubsection~\ref{subsubsec2}).

\section{Tables}\label{sec5}

Tables can be inserted via the normal table and tabular environment. To put
footnotes inside tables you should use \verb+\footnotetext[]{...}+ tag.
The footnote appears just below the table itself (refer Tables~\ref{tab1} and \ref{tab2}). 
For the corresponding footnotemark use \verb+\footnotemark[...]+

\begin{table}[h]
\begin{center}
\begin{minipage}{174pt}
\caption{Caption text}\label{tab1}%
\begin{tabular}{@{}llll@{}}
\toprule
Column 1 & Column 2  & Column 3 & Column 4\\
\midrule
row 1    & data 1   & data 2  & data 3  \\
row 2    & data 4   & data 5\footnotemark[1]  & data 6  \\
row 3    & data 7   & data 8  & data 9\footnotemark[2]  \\
\botrule
\end{tabular}
\footnotetext{Source: This is an example of table footnote. This is an example of table footnote.}
\footnotetext[1]{Example for a first table footnote. This is an example of table footnote.}
\footnotetext[2]{Example for a second table footnote. This is an example of table footnote.}
\end{minipage}
\end{center}
\end{table}

\noindent
The input format for the above table is as follows:

%%=============================================%%
%% For presentation purpose, we have included  %%
%% \bigskip command. please ignore this.       %%
%%=============================================%%
\bigskip
\begin{verbatim}
\begin{table}[<placement-specifier>]
\begin{center}
\begin{minipage}{<preferred-table-width>}
\caption{<table-caption>}\label{<table-label>}%
\begin{tabular}{@{}llll@{}}
\toprule
Column 1 & Column 2 & Column 3 & Column 4\\
\midrule
row 1 & data 1 & data 2	 & data 3 \\
row 2 & data 4 & data 5\footnotemark[1] & data 6 \\
row 3 & data 7 & data 8	 & data 9\footnotemark[2]\\
\botrule
\end{tabular}
\footnotetext{Source: This is an example of table footnote. 
This is an example of table footnote.}
\footnotetext[1]{Example for a first table footnote.
This is an example of table footnote.}
\footnotetext[2]{Example for a second table footnote. 
This is an example of table footnote.}
\end{minipage}
\end{center}
\end{table}
\end{verbatim}
\bigskip
%%=============================================%%
%% For presentation purpose, we have included  %%
%% \bigskip command. please ignore this.       %%
%%=============================================%%

\begin{table}[h]
\begin{center}
\begin{minipage}{\textwidth}
\caption{Example of a lengthy table which is set to full textwidth}\label{tab2}
\begin{tabular*}{\textwidth}{@{\extracolsep{\fill}}lcccccc@{\extracolsep{\fill}}}
\toprule%
& \multicolumn{3}{@{}c@{}}{Element 1\footnotemark[1]} & \multicolumn{3}{@{}c@{}}{Element 2\footnotemark[2]} \\\cmidrule{2-4}\cmidrule{5-7}%
Project & Energy & $\sigma_{calc}$ & $\sigma_{expt}$ & Energy & $\sigma_{calc}$ & $\sigma_{expt}$ \\
\midrule
Element 3  & 990 A & 1168 & $1547\pm12$ & 780 A & 1166 & $1239\pm100$\\
Element 4  & 500 A & 961  & $922\pm10$  & 900 A & 1268 & $1092\pm40$\\
\botrule
\end{tabular*}
\footnotetext{Note: This is an example of table footnote. This is an example of table footnote this is an example of table footnote this is an example of~table footnote this is an example of table footnote.}
\footnotetext[1]{Example for a first table footnote.}
\footnotetext[2]{Example for a second table footnote.}
\end{minipage}
\end{center}
\end{table}

In case of double column layout, tables which do not fit in single column width should be set to full text width. For this, you need to use \verb+\begin{table*}+ \verb+...+ \verb+\end{table*}+ instead of \verb+\begin{table}+ \verb+...+ \verb+\end{table}+ environment. Lengthy tables which do not fit in textwidth should be set as rotated table. For this, you need to use \verb+\begin{sidewaystable}+ \verb+...+ \verb+\end{sidewaystable}+ instead of \verb+\begin{table*}+ \verb+...+ \verb+\end{table*}+ environment. This environment puts tables rotated to single column width. For tables rotated to double column width, use \verb+\begin{sidewaystable*}+ \verb+...+ \verb+\end{sidewaystable*}+.

\begin{sidewaystable}
\sidewaystablefn%
\begin{center}
\begin{minipage}{\textheight}
\caption{Tables which are too long to fit, should be written using the ``sidewaystable'' environment as shown here}\label{tab3}
\begin{tabular*}{\textheight}{@{\extracolsep{\fill}}lcccccc@{\extracolsep{\fill}}}
\toprule%
& \multicolumn{3}{@{}c@{}}{Element 1\footnotemark[1]}& \multicolumn{3}{@{}c@{}}{Element\footnotemark[2]} \\\cmidrule{2-4}\cmidrule{5-7}%
Projectile & Energy	& $\sigma_{calc}$ & $\sigma_{expt}$ & Energy & $\sigma_{calc}$ & $\sigma_{expt}$ \\
\midrule
Element 3 & 990 A & 1168 & $1547\pm12$ & 780 A & 1166 & $1239\pm100$ \\
Element 4 & 500 A & 961  & $922\pm10$  & 900 A & 1268 & $1092\pm40$ \\
Element 5 & 990 A & 1168 & $1547\pm12$ & 780 A & 1166 & $1239\pm100$ \\
Element 6 & 500 A & 961  & $922\pm10$  & 900 A & 1268 & $1092\pm40$ \\
\botrule
\end{tabular*}
\footnotetext{Note: This is an example of table footnote this is an example of table footnote this is an example of table footnote this is an example of~table footnote this is an example of table footnote.}
\footnotetext[1]{This is an example of table footnote.}
\end{minipage}
\end{center}
\end{sidewaystable}

\section{Figures}\label{sec6}

As per the \LaTeX\ standards you need to use eps images for \LaTeX\ compilation and \verb+pdf/jpg/png+ images for \verb+PDFLaTeX+ compilation. This is one of the major difference between \LaTeX\ and \verb+PDFLaTeX+. Each image should be from a single input .eps/vector image file. Avoid using subfigures. The command for inserting images for \LaTeX\ and \verb+PDFLaTeX+ can be generalized. The package used to insert images in \verb+LaTeX/PDFLaTeX+ is the graphicx package. Figures can be inserted via the normal figure environment as shown in the below example:

%%=============================================%%
%% For presentation purpose, we have included  %%
%% \bigskip command. please ignore this.       %%
%%=============================================%%
\bigskip
\begin{verbatim}
\begin{figure}[<placement-specifier>]
\centering
\includegraphics{<eps-file>}
\caption{<figure-caption>}\label{<figure-label>}
\end{figure}
\end{verbatim}
\bigskip
%%=============================================%%
%% For presentation purpose, we have included  %%
%% \bigskip command. please ignore this.       %%
%%=============================================%%

\begin{figure}[h]%
\centering
\includegraphics[width=0.9\textwidth]{fig.eps}
\caption{This is a widefig. This is an example of long caption this is an example of long caption  this is an example of long caption this is an example of long caption}\label{fig1}
\end{figure}

In case of double column layout, the above format puts figure captions/images to single column width. To get spanned images, we need to provide \verb+\begin{figure*}+ \verb+...+ \verb+\end{figure*}+.

For sample purpose, we have included the width of images in the optional argument of \verb+\includegraphics+ tag. Please ignore this.

\section{Algorithms, Program codes and Listings}\label{sec7}

Packages \verb+algorithm+, \verb+algorithmicx+ and \verb+algpseudocode+ are used for setting algorithms in \LaTeX\ using the format:

%%=============================================%%
%% For presentation purpose, we have included  %%
%% \bigskip command. please ignore this.       %%
%%=============================================%%
\bigskip
\begin{verbatim}
\begin{algorithm}
\caption{<alg-caption>}\label{<alg-label>}
\begin{algorithmic}[1]
. . .
\end{algorithmic}
\end{algorithm}
\end{verbatim}
\bigskip
%%=============================================%%
%% For presentation purpose, we have included  %%
%% \bigskip command. please ignore this.       %%
%%=============================================%%

You may refer above listed package documentations for more details before setting \verb+algorithm+ environment. For program codes, the ``program'' package is required and the command to be used is \verb+\begin{program}+ \verb+...+ \verb+\end{program}+. A fast exponentiation procedure:

\begin{program}
\BEGIN \\ %
  \FOR i:=1 \TO 10 \STEP 1 \DO
     |expt|(2,i); \\ |newline|() \OD %
\rcomment{Comments will be set flush to the right margin}
\WHERE
\PROC |expt|(x,n) \BODY
          z:=1;
          \DO \IF n=0 \THEN \EXIT \FI;
             \DO \IF |odd|(n) \THEN \EXIT \FI;
\COMMENT{This is a comment statement};
                n:=n/2; x:=x*x \OD;
             \{ n>0 \};
             n:=n-1; z:=z*x \OD;
          |print|(z) \ENDPROC
\END
\end{program}

\begin{algorithm}
\caption{Calculate $y = x^n$}\label{algo1}
\begin{algorithmic}[1]
\Require $n \geq 0 \vee x \neq 0$
\Ensure $y = x^n$ 
\State $y \Leftarrow 1$
\If{$n < 0$}\label{algln2}
        \State $X \Leftarrow 1 / x$
        \State $N \Leftarrow -n$
\Else
        \State $X \Leftarrow x$
        \State $N \Leftarrow n$
\EndIf
\While{$N \neq 0$}
        \If{$N$ is even}
            \State $X \Leftarrow X \times X$
            \State $N \Leftarrow N / 2$
        \Else[$N$ is odd]
            \State $y \Leftarrow y \times X$
            \State $N \Leftarrow N - 1$
        \EndIf
\EndWhile
\end{algorithmic}
\end{algorithm}
\bigskip
%%=============================================%%
%% For presentation purpose, we have included  %%
%% \bigskip command. please ignore this.       %%
%%=============================================%%

Similarly, for \verb+listings+, use the \verb+listings+ package. \verb+\begin{lstlisting}+ \verb+...+ \verb+\end{lstlisting}+ is used to set environments similar to \verb+verbatim+ environment. Refer to the \verb+lstlisting+ package documentation for more details.

%%=============================================%%
%% For presentation purpose, we have included  %%
%% \bigskip command. please ignore this.       %%
%%=============================================%%
\bigskip
\begin{minipage}{\hsize}%
\lstset{frame=single,framexleftmargin=-1pt,framexrightmargin=-17pt,framesep=12pt,linewidth=0.98\textwidth,language=pascal}% Set your language (you can change the language for each code-block optionally)
%%% Start your code-block
\begin{lstlisting}
for i:=maxint to 0 do
begin
{ do nothing }
end;
Write('Case insensitive ');
Write('Pascal keywords.');
\end{lstlisting}
\end{minipage}

\section{Cross referencing}\label{sec8}

Environments such as figure, table, equation and align can have a label
declared via the \verb+\label{#label}+ command. For figures and table
environments use the \verb+\label{}+ command inside or just
below the \verb+\caption{}+ command. You can then use the
\verb+\ref{#label}+ command to cross-reference them. As an example, consider
the label declared for Figure~\ref{fig1} which is
\verb+\label{fig1}+. To cross-reference it, use the command 
\verb+Figure \ref{fig1}+, for which it comes up as
``Figure~\ref{fig1}''.

To reference line numbers in an algorithm, consider the label declared for the line number 2 of Algorithm~\ref{algo1} is \verb+\label{algln2}+. To cross-reference it, use the command \verb+\ref{algln2}+ for which it comes up as line~\ref{algln2} of Algorithm~\ref{algo1}.

\subsection{Details on reference citations}\label{subsec7}

Standard \LaTeX\ permits only numerical citations. To support both numerical and author-year citations this template uses \verb+natbib+ \LaTeX\ package. For style guidance please refer to the template user manual.

Here is an example for \verb+\cite{...}+: \cite{bib1}. Another example for \verb+\citep{...}+: \citep{bib2}. For author-year citation mode, \verb+\cite{...}+ prints Jones et al. (1990) and \verb+\citep{...}+ prints (Jones et al., 1990).

All cited bib entries are printed at the end of this article: \cite{bib3}, \cite{bib4}, \cite{bib5}, \cite{bib6}, \cite{bib7}, \cite{bib8}, \cite{bib9}, \cite{bib10}, \cite{bib11} and \cite{bib12}.

\section{Examples for theorem like environments}\label{sec10}

>>>>

%%\documentclass[sn-basic]{sn-jnl}% Basic Springer Nature Reference Style/Chemistry Reference Style
%%\documentclass[sn-mathphys]{sn-jnl}% Math and Physical Sciences Reference Style
%%\documentclass[sn-aps]{sn-jnl}% American Physical Society (APS) Reference Style
%%\documentclass[sn-vancouver]{sn-jnl}% Vancouver Reference Style
\documentclass[pdflatex,sn-apa]{sn-jnl}% APA Reference Style
%%\documentclass[sn-chicago]{sn-jnl}% Chicago-based Humanities Reference Style
%%\documentclass[sn-standardnature]{sn-jnl}% Standard Nature Portfolio Reference Style
%%\documentclass[default]{sn-jnl}% Default
%%\documentclass[default,iicol]{sn-jnl}% Default with double column layout

%%%% Standard Packages
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{csquotes}
%%<additional latex packages if required can be included here>
%%%%

@@

@@

\title[RUNNING HEAD: Similarity, Contrast, and Representativeness in Categorization]{Creating Something Different: Similarity, Contrast, and Representativeness in Categorization}

@@

\author*[1]{\fnm{Joseph L.} \sur{Austerweil}}\email{austerweil@wisc.edu}
\equalcont{These authors contributed equally to this work.}

\author[2]{\fnm{Shi Xian} \sur{Liew}}\email{shixianliew@gmail.com}
\equalcont{These authors contributed equally to this work.}

\author[1]{\fnm{Nolan} \sur{Conaway}}\email{nolanbconaway@gmail.com}

\author[3]{\fnm{Kenneth J.} \sur{Kurtz}}\email{kkurtz@binghamton.edu}

\affil*[1]{\orgdiv{Department of Psychology}, \orgname{University of Wisconsin-Madison}, \orgaddress{\street{1202 West Johnson Street}, \city{Madison}, \postcode{53706}, \state{WI}, \country{USA}}}

\affil[2]{\orgdiv{School of Psychology}, \orgname{University of South New Wales}, \orgaddress{\street{Mathews Building}, \city{Sydney}, \postcode{2052}, \state{NSW}, \country{Australia}}}

\affil[3]{\orgdiv{Department of Psychology }\orgname{SUNY Binghamton} \orgaddress{\city{Binghamton}, \postcode{{12042}}, \state{NY} \country{USA}}}

@@

\abstract{The ability to generate new concepts and ideas is among the most fascinating aspects of human cognition, but we do not have a strong understanding of the
cognitive processes and representations underlying concept generation. In this
paper, we study the generation of new categories using the computational and
behavioral toolkit of traditional artificial category learning. Previous work in
this domain has focused on how the statistical structure of known categories
generalizes to generated categories, overlooking whether (and if so, how)
contrast between the known and generated categories is a factor. We report three
experiments demonstrating that contrast between what is known and what is
created is of fundamental importance for categorization. We propose two novel
approaches to modeling category contrast: one focused on exemplar dissimilarity
and another on the representativeness heuristic. Our experiments and
computational analyses demonstrate that both models capture different aspects of
contrast's role in categorization.}

@@

\keywords{categorization, concepts, category learning, generation,
computational modeling, exemplar models, representativeness}

@@

\section{Declarations Section}
\subsection{Funding}
The studies were funded by the Wisconsin Alumni Research Foundation through University of Wisconsin-Madison Office of the Vice Chancellor for Graduate Education.

\subsection{Conflict of Interests}
No authors have any conflicts of interests to declare.

\subsection{Availability of Data,  Materials, and Code }
All experiment materials, anonymized raw data, and code are available at \url{https://github.com/AusterweilLab/generating_categories}.

\subsection{Authors Contributions}
The first draft of the manuscript was primarily written by JLA and SXL. All authors contributed to writing the manuscript substantially by editing and/or commenting on the manuscript. The computational models, and Experiments 1 and 2 were designed by JLA, NC, and KJK. Experiments 1 and 2 were implemented and conducted by NC and JLA. Experiment 3 was designed by JLA, SXL, and KJK. It was implemented and conducted by SXL and JLA. Model-based and other analyses were done by NC, SXL, and JLA.

\subsection{Ethical Approval}
All human subject experiments were originally approved by the Social Behavioral and Education Institutional Review Board (now named the Minimal Risk Research Institutional Review Board) at the University of Wisconsin-Madison (protocol number 2016-1011).

\subsection{Consent to Participate}
All participants were provided informed consent before participating. Given the minimal risk provided by the study, it was determined that an unintentional breach of confidentiality would be the most likely potential risk and so, no identifying information was stored.

\subsection{Consent to Publish}
Not Applicable.

\newpage

\section{Introduction}\label{sec1}
Creating new ideas is one of the most fascinating and important
human capabilities. For example, cell phones and smart phones are newly created
categories of objects that young and old adults learned and have become reliant
on in the last few decades. Computational researchers have identified
generating novel objects and objectives as one of the defining, and most
difficult to formalize, characteristics of intelligent life
\citep{lake2017,lehman2011,taylor2016}. Recently, generative deep learning methods, such as stable diffusion \cite{rombach2022} and ChatGPT \cite{chatgpt2022}, have captivated researchers and the overall public. Their power to generate novel images and text is astonishing. They have been reported to win an art contest \cite{roose2022}, and to pass a major graduate medical and business school examinations \cite{kung2022,terwiesch2023}. What are the major principles underlying category generation?

Despite recent successes in generation from machine learning methods, the cognitive mechanisms that enable people to innovate are not well understood and
are understudied by scientists. In part, this is due to the inherent difficulty
of designing and conducting experiments that test these capabilities. How would
a scientist devise an experiment that induces a participant to create 
new categories that are as interesting as cell phones and smart phones? Devising such an experiment is still beyond our capability. Instead, we take a step towards this goal, and investigate the fundamental constraints and expectations that people have when they generate new categories.

% we do have tools to study the fundamental constraints on category . As a
%step towards these goals, in this article we investigate the expectations that
%people have when they generate new categories using computational
%models and behavioral experiments.

Creating new concepts are, however, not altogether different from
the types of behaviors typically studied in cognitive psychology laboratories.
In particular, generating a member of a novel class can be considered a `special
case' application of existing category knowledge
\citep{kemp2014taxonomy,kurtz2015human}. Research in categorization typically
focuses on what properties of a category affect human learning of an object's
category given its features \citep{shepard1961learning,kurtz2013human}, or the
prediction of an object's unobserved features given some of its other features
and/or its category \citep{markman2003category}. The generation of members of a
new category consists of inferring {\em all} features for a {\em novel} category
label. Thus, we can make progress formalizing the processes involved in category
generation by extending theories of categorization to this case. This will also serve to test the breadth of the explanatory power of categorization models to tasks beyond typical category learning.

Previous work in category learning and generation has established that people
are highly sensitive to the structural properties of categories, such as
correlations between the features of category members and the relation between
items within the same category and those in different categories
\citep{roschmervis1975,regier2007,shepard1961learning,smith1995creative}.\footnote{It
  is worth noting that sensitivity to feature correlations is not universal and
  can be dependent on factors such as the salience of correlations and task
  demands \citep[e.g., see][]{chin2002effect,malt1984correlated,murphy2004big}.}
Inspired by this work, previous research on the topic of category generation has
explored a similar principle: People tend to create new categories that have
similar {\em statistical regularities} as previously learned categories
\citep{jern2013probabilistic,ward1994structured}. These statistical regularities
reflect general structural patterns across old and new categories and are not
restricted to strictly quantitatively similar properties.

Although statistical regularity is an important characteristic of generating new
categories, it cannot be the only one. Taken to the extreme, the best ``new''
category in terms of having the same statistical regularities to other
categories would be identical to a known category that is representative of the
domain (and thus, not new at all). Contrast from other known categories should
play a role. Indeed, scientists in other fields, such as marketing
\citep{berger16} and sociology \citep{rogers03}, highlight the critical
importance of contrast in the creation of new ideas.

To successfully generate something novel, what is generated must be different
from what is already known. This fundamental constraint, ``being different'', or
contrasting from other categories in the relevant domain, is the focus of our
work. Intuitively, the influence of contrast may be expected in a number of
real-life scenarios involving the generation of new categories. For instance, a
college student who has been eating instant noodles for a week may be more
inclined to seek out something healthier and less carbohydrate-heavy (e.g., a salad) compared to something similar (a different flavor of instant
noodles). An actor who is dissatisfied with being typecast as a villain may seek
out roles as heroes as opposed to playing other antagonistic characters. An
especially compelling example can be seen in \cite{askin2017} who found that new
songs that were more different to their peers were generally more popular than
new songs that were similar to others. It appears that once there is the
motivation to generate something new, there is an associated motivation to
generate and observe something \emph{different}.

Although implicitly assumed in some work, this constraint has been
overlooked in previous research: To our knowledge, there has not been any
systematic investigation addressing how generated categories {\em differ} from
what is already known. Although the idea of category contrast is discussed
throughout the categorization literature, and extends to a variety of other
fields \citep[e.g., color;][]{regier2007}, the idea that a new category should
be ``different'' is vague, as there are many ways it could be different from a
previously observed category.

We examine two different definitions of how categories should differ from one
another: exemplar dissimilarity and representativeness of the alternative
category. To formalize the first technique, we build on the largely successful
exemplar modeling framework
\citep{medin1978context,nosofsky1984choice,nosofsky1986attention}. In doing so,
we propose a novel exemplar model of category generation, {\em Producing Alike
  and Contrasting Knowledge using Exemplar Representations} (PACKER),
formalizing how new categories should differ from previous categories. It
embodies the exemplar dissimilarity principle by incorporating a factor that
repulses members of opposite categories from one another.

The second hypothesis for how contrast might affect categorization is contrast
as {\em representativeness}. The representativeness heuristic
\citep{kahneman1972subjective} states that people make judgments regarding an
outcome based on how representative it is of the evidence given in the current
context. The heuristic is a powerful theoretical construct that has been used to
capture a range of complex patterns of human judgments
\citep{kahneman73,tversky74,tversky83}, especially those that deviate from
normative theory (as given by a straightforward application of Bayes' rule). For
example, the coinflip sequence TTHTH is perceived to be more random (due to it
being representative of a coinflip sequence) than TTTTT (despite both having the
same probability of having been generated from a fair coin). Although there is a
healthy debate as to whether perceived randomness {\em really} deviates from
normative theory \citep{griffiths18,hahn09}, the representativeness heuristic
remains one of the main explanations of human judgments \citep{reimers18}. To
formalize the hypothesis that people expect a new category to be representative
of the opposite of the current category, or contrast as {\em
  representativeness}, we use the Bayesian formulation of representativeness
\citep{tenenbaum01}, which has been used successfully to capture human
performance in color categorization \citep{abbott16}, image category learning
\citep{abbott11}, and language grammar learning \citep{rafferty10}. We do so by
extending the state-of-the-art category generation model
\citep{jern2013probabilistic} and find that human category generation can
also be captured as generating representative, rather than probable, samples.

The outline of the article is as follows. First we describe previous
computational formalizations of theories of category generation and empirical
work investigating them. We then present two hypotheses for how contrast might
affect categorization and formalize them in computational models. The first is a
novel exemplar model, which is designed to generate categories that
systematically differ from existing categories in the domain. The second assumes
that the goal of category generation is to create representative samples of the
opposite of the observed categories. We present two experiments demonstrating
strong and systematic effects of category contrast on concept generation, and we
qualitatively and quantitatively analyze the behavioral results and the performance of each model in
capturing human category generation. A third experiment is performed to
highlight the differences between the two contrast models. We conclude with a
discussion of the implications of our results for categorization and directions
for future work.

\section{Prior work}

Much of what we know about concept generation and contrast comes from the
foundational literature on creative cognition. In a series of reports, Ward and
colleagues
\citep{ward1995s,ward1994structured,marsh1999inadvertent,ward2002role,smith1993constraining}
established that category generation is highly constrained by prior knowledge:
Generated categories tend to consist of features observed in known categories,
and they tend to exhibit the distributional properties found in known
categories. In a seminal study, \cite{ward1994structured} asked participants to
generate new species of alien animals by drawing and describing members of the
species. People tended to generate species with the same features as on Earth
(e.g., eyes, legs, wings), and possessing the same feature correlations as on
Earth (e.g., feathers co-occur with wings). Likewise, aliens drawn from the same
species tended to share more features with one another compared to members of
opposite species.

The broader set of observations made by Ward and colleagues provide a great deal
of insight into the role of prior knowledge in constraining category generation.
Much of the work from this area
\citep[e.g.,][]{smith1993constraining,marsh1999inadvertent} focuses on how
information provided to participants (such as an example of a species generated
by other participants) can drastically diminish the difference of a new category
from pre-existing categories. Theoretical accounts of these effects have
primarily been grounded within the categorization literature. For example, the
predominant ``Path of Least Resistance'' account
\citep[see][]{ward1994structured,ward1995s,ward2002role} proposes that, when
generating a new species of animal, people retrieve from memory a known
subcategory of animals (e.g., {\em bird}, {\em dog}, {\em horse}), and simply
change some of the features to make something new. People are thought to change
only features that are not characteristic of the retrieved category (e.g., if
{\em bird} was retrieved, the presence of {\em wings} would not change, but {\em color} might). This theory incorporates elements of the highly influential
basic-level categories framework \citep{rosch1975cognitive,rosch1976basic}, as
well as the exemplar view \citep{medin1978context,brooks1978nonanalytic}. While
this work has been incredibly useful in providing a conceptual sketch of
generation theories, its qualitative nature and the hand-drawn responses used in
the experiments paradigms precludes the development of formal approaches that
can be used to test them in a fine-grained manner. One other relevant study of note is \cite{lake2015}. In one part of their experiments, they had participants generate novel alphabets. While their work is fascinating, the model is domain-specific (representing orthography using a grammar over the parameters of B\'{e}zier curves) and the results are difficult to generalize beyond that domain.

\cite{jern2013probabilistic} showed that concept generation could be
studied in a more controlled manner through the well-developed methods of an
artificial categorization paradigm \citep[see][for a review]{kurtz2015human}. In
Experiments 3 and 4 of their article, participants were exposed to members of
experimenter-defined categories of "crystals" varying in size, hue, and
saturation. Following a training phase during which the experimenter-defined
categories were learned, participants were asked to generate novel categories of
crystals. In a finding mirroring \cite{ward1994structured},
\cite{jern2013probabilistic} found that participants generated categories with
the same distributional properties as the experimenter-defined categories. For
example, after learning categories with a positive correlation between the size
and saturation features (larger sized crystals were more saturated),
participants generated novel categories with the same positive correlation. By
replicating \cite{ward1994structured}, they demonstrated that category generation
can be studied in a well-known and highly controlled experimental paradigm.

The authors evaluated the predictions of several formal models on their data.
Most notably, they showed that a hierarchical Bayesian model provided the
strongest account of their results. Their model views observed examples as
samples from an underlying category distribution, describing the location of the
category in the space, as well as how it varies along each feature. In turn,
each category is viewed as a sample from an underlying {\em domain}
distribution, specifying distributional commonalities among the observed
categories. Generated categories are thought to stem from the same domain
distribution as observed categories, thus the distributional properties of
observed categories will be preserved within the generated category.

\cite{jern2013probabilistic} additionally tested a ``copy-and-tweak'' model that
broadly resembles the earlier ``Path of Least Resistance'' account. The core
proposal is that participants generate new items by copying stored examples from
memory and tweaking them to generate something new. The copy-and-tweak model
differs from the Path of Least Resistance account in that it notably omits the
hierarchical organization of categories, as well as selectivity to which
features are changed \citep[both of which are factors in the Path of Least
Resistance account;][]{ward2002role}. Instead, their copy-and-tweak model
corresponds to a direct exemplar-similarity approach
\citep[e.g.,][]{nosofsky1984choice,nosofsky1986attention}, generating new items
according to their similarity to known members of the target category. The
copy-and-tweak model provided a poor account of the results of \cite{jern2013probabilistic}.

It is worth pointing out that the early observation of regularities in
distributional properties across categories is not confined to the work of Ward
and colleagues. Most notably, \citet{thomas98} trained participants to classify
circles of different sizes, each with a radial line of varying orientation, into
two categories. During a surprise prediction phase, participants were asked to
generate a value for a missing feature for a certain category exemplar given a
value on the other feature. Their results revealed that exemplars tended to
share the same feature correlations as exemplars from previously learned
categories. Interestingly, this effect was not consistently observed when the
given values of a feature fell outside the range of learned values for that
particular category. For instance, if the learned Category A exemplars were
generally 8 cm to 10 cm in radius and negatively correlated with the angle of
the radial line, participants generally produced radial line orientations that
were also negatively correlated with the exemplar size when the given size was
between 8 cm and 10 cm. However, when the given size fell outside that range for
that category (e.g., 14 cm for a Category A exemplar), some participants
consistently produced radial line orientations that were not negatively
correlated with the size and that also placed the exemplars in a significantly
different location in the feature space. These results suggest that when tasked
to produce exemplars that fall outside what was previously learned, people show
some tendency to generate exemplars that are different (both distributionally
and spatially) to what was learned. We explore this idea of contrast
more deeply in the following section.

\section{And Now for Something (Completely) Different: The Role of Contrast}

%Prior work on category generation has explored only one factor: distributional
%correspondences between learned and generated categories. As a result, most of
%the computational, theoretical, and empirical efforts have been focused on
%explaining those effects. In this paper, we investigate another important
%factor: category contrast. To generate a novel concept, individuals must produce
%something that is in some capacity {\em different} from what they already know.
%As a consequence, we propose that contrast should be a primary constraint that
%underlies people's expectations about how a set of categories are organized:
%Categories should be different from each other.
